A multi-species benchmark for training and validating large scale mass spectrometry proteomics machine learning models

### 2024-07-17 doc/nine-species/nine-species.tex
Small updates

\begin{table}
\scriptsize
\centering
\begin{tabular}{lrrrrrr}
\hline
Species & \multicolumn{2}{c}{Overlap} & \multicolumn{2}{c}{Main} & \multicolumn{2}{c}{Balanced} \\
& PSMs & Peptides  & PSMs & Peptides  & PSMs & Peptides  \\
\hline
\textit{Vigna mungo} & & & 108514 & 12001 \\
\textit{Mus musculus} & & & 25541 & 5899 \\
\textit{Methanosarcina mazei} & & & 267333 & 15925 \\
\textit{Bacillus subtilis} & & & 1358337 & 30786 \\
\textit{Candidatus endoloripes} & & & 82290 & 8392 \\
\textit{Solanum lycopersicum} & & & 178413 & 49745 \\
\textit{Saccharomyces-cerevisiae} & & & 585593 & 19720 \\
\textit{Apis mellifera} & & & 194281 & 21559 \\
\textit{H. sapiens} & & & 44604 & 11289 \\
\hline
Total & & & 2,844,906 & 175,316 \\
\end{tabular}
\caption{{\bf Three versions of the nine-species benchmark.}
  The ``overlap'' dataset contains peptides that are associated with more than one species.
  The ``main'' dataset \fixme{XXX}
}
\label{tab:benchmark}
\end{table}

@@

We assigned peptide labels to spectra using the Tide search engine followed by post-processing with Percolator.
In creating the Tide index, we specified specified Cys carbamidomethylation as a static modification and allowed for the following variable modifications: Met oxidation, Asn deamidation, Gln deamidation, N-term acetylation, N-term carbamylation, N-term NH$_{3}$ loss, and the combination of N-term carbamylation and NH$_{3}$ loss by using the tide-index options {\tt --mods-spec 1M+15.994915,\allowbreak 1N+0.984016,\allowbreak 1Q+0.984016 --nterm-peptide-mods-spec 1X+42.010565,\allowbreak 1X+43.005814,\allowbreak 1X-17.026549,\allowbreak 1X+25.980265 --max-mods 3}.
Note that one of the nine experiments (\textit{Mus musculus}) was performed using SILAC labeling, but we searched without SILAC modifications and hence include in the benchmark only PSMs from unlabeled peptides.
Tide automatically added to each index a shuffled decoy peptide corresponding to each target peptide.
Thereafter, each MGF file was searched against the corresponding index using the precursor window size and fragment bin tolerance specified in the original study (Table~\ref{tab:benchmark}).
The search engine employed XCorr scoring with Tailor calibration \cite{sulimov2020tailor}, and we allowed for 1 isotope error in the selection of candidate peptides.
All search results were then analyzed jointly per species using the Crux implementation of Percolator, with default parameters.
For the benchmark, we retained all PSMs with Percolator q value $<$ 0.01.
We identified 13 MGF files with fewer than 100 accepted PSMs, and we eliminated all of these PSMs from the benchmark.
The resulting benchmark (``overlap'') contains \fixme{XXX} million PSMs drawn from 343 RAW files.

@@

To avoid train/test leakage, we post-processed the PSMs to eliminate peptides that are shared between species.
Among the 229,984 unique peptides, we identified 3797 (1.7\%) that occur in more than one species.
For each such peptide, we selected one of the associated species at random and then eliminated all PSMs containing that peptide in other species.
Note that when identifying shared peptides between species, we considered all modified forms of a given peptide sequence to be the same.
Hence, if a given peptide appears in more than one species, then that peptide, including all its modified forms, is randomly assigned to a single species and eliminated from the others.
The final, non-redundant benchmark dataset (``main'') consists of 2.8 million PSMs corresponding to \fixme{XXX} peptides.

@@

At this stage, the benchmark was quite imbalanced, in the sense that some species had a much larger number of associated PSMs.
We therefore used a random downsampling procedure to produce a benchmark that is more evenly balanced across species.
Among the nine species, the one with the fewest PSMs is \textit{Mus musculus}, with 25,541.
Downsampling all of the other eight species to have 25,000 PSMs would reduce the size of the dataset from 2.8 million PSMs to 225,000---a reduction of 92\%.
To avoid producing such a small dataset, we therefore opted to downsample each dataset to approximately 100,000 PSMs.
This approach yields a slight imbalance, because three species have fewer than 100,000 PSMs (44,604 for \textit{H.\ sapiens} and 82,290 for \textit{Candidatus endoloripes}), while retaining a larger percentage of the original data.
Our downsampling procedure involved randomly permuting the order of the MGF files for each species and then selecting the files in order until at least 100,000 PSMs have been accepted.
The final, balanced benchmark dataset consists of \fixme{XXX} PSMs from \fixme{XXX} peptides.

@@

The dataset contains files resulting from various stages of the benchmark:
\begin{itemize}
\item Spectrum files in MGF format, produced by msconvert.
\item Reference proteome files in FASTA format, downloaded from Uniprot.
\item Search results files for both targets and decoys, in tab-delimited format, produced by Tide.
\item PSM-level Percolator results files for targets, in tab-delimited format.
\item Annotated MGF files for all three versions of the benchmark (overlap, main, and balanced).
\end{itemize}
Also included are log files for the steps of the analysis pipeline carried out using Crux \cite{park2008rapid} (Tide indexing, Tide search, and Percolator).

>>>>

\begin{table}
\scriptsize
\centering
\begin{tabular}{lrrrrrr}
\hline
Species & \multicolumn{2}{c}{Main} & \multicolumn{2}{c}{Balanced} \\
& PSMs & Peptides  & PSMs & Peptides  \\
\hline
\textit{Vigna mungo} & & & 108514 & 12001 \\
\textit{Mus musculus} & & & 25541 & 5899 \\
\textit{Methanosarcina mazei} & & & 267333 & 15925 \\
\textit{Bacillus subtilis} & & & 1358337 & 30786 \\
\textit{Candidatus endoloripes} & & & 82290 & 8392 \\
\textit{Solanum lycopersicum} & & & 178413 & 49745 \\
\textit{Saccharomyces-cerevisiae} & & & 585593 & 19720 \\
\textit{Apis mellifera} & & & 194281 & 21559 \\
\textit{H. sapiens} & & & 44604 & 11289 \\
\hline
Total & & & 2,844,906 & 175,316 \\
\end{tabular}
\caption{{\bf Two versions of the nine-species benchmark.}
  The balanced dataset eliminates some data in order to approximately balance the amount of data associated with each species.
}
\label{tab:benchmark}
\end{table}

@@

We assigned peptide labels to spectra using the Tide search engine followed by post-processing with Percolator.
In creating the Tide index, we specified specified Cys carbamidomethylation as a static modification and allowed for the following variable modifications: Met oxidation, Asn deamidation, Gln deamidation, N-term acetylation, N-term carbamylation, N-term NH$_{3}$ loss, and the combination of N-term carbamylation and NH$_{3}$ loss by using the tide-index options {\tt --mods-spec 1M+15.994915,\allowbreak 1N+0.984016,\allowbreak 1Q+0.984016 --nterm-peptide-mods-spec 1X+42.010565,\allowbreak 1X+43.005814,\allowbreak 1X-17.026549,\allowbreak 1X+25.980265 --max-mods 3}.
Note that one of the nine experiments (\textit{Mus musculus}) was performed using SILAC labeling, but we searched without SILAC modifications and hence include in the benchmark only PSMs from unlabeled peptides.
Tide automatically added to each index a shuffled decoy peptide corresponding to each target peptide.
Thereafter, each MGF file was searched against the corresponding index using the precursor window size and fragment bin tolerance specified in the original study (Table~\ref{tab:benchmark}).
The search engine employed XCorr scoring with Tailor calibration \cite{sulimov2020tailor}, and we allowed for 1 isotope error in the selection of candidate peptides.
All search results were then analyzed jointly per species using the Crux implementation of Percolator, with default parameters.
For the benchmark, we retained all PSMs with Percolator q value $<$ 0.01.
We identified 13 MGF files with fewer than 100 accepted PSMs, and we eliminated all of these PSMs from the benchmark.
At this point in the processing pipeline, the dataset contains 2,898,611 annotated spectra (PSMs)  drawn from 339 RAW files and associated with 179,256 distinct peptides.

@@

To avoid train/test leakage, we post-processed the PSMs to eliminate peptides that are shared between species.
Among the 229,984 unique peptides, we identified 3797 (1.7\%) that occur in more than one species.
For each such peptide, we selected one of the associated species at random and then eliminated all PSMs containing that peptide in other species.
Note that when identifying shared peptides between species, we considered all modified forms of a given peptide sequence to be the same, and we converted all isoleucines to leucines.
Hence, if a given peptide appears in more than one species, then that peptide, including all its modified forms, is randomly assigned to a single species and eliminated from the others.
The final, non-redundant benchmark dataset (``main'') consists of 2,843,631 PSMs corresponding to 258,141 distinct peptides.

@@

At this stage, the benchmark was quite imbalanced, in the sense that some species had a much larger number of associated PSMs.
We therefore used a random downsampling procedure to produce a benchmark that is more evenly balanced across species.
Among the nine species, the one with the fewest PSMs is \textit{Mus musculus}, with 25,541.
Downsampling all of the other eight species to have 25,000 PSMs would reduce the size of the dataset from 2.8 million PSMs to 225,000---a reduction of 92\%.
To avoid producing such a small dataset, we therefore opted to downsample each dataset to approximately 100,000 PSMs.
This approach yields a slight imbalance, because three species have fewer than 100,000 PSMs (44,604 for \textit{H.\ sapiens} and 82,290 for \textit{Candidatus endoloripes}), while retaining a larger percentage of the original data.
Our downsampling procedure involved randomly permuting the order of the MGF files for each species and then selecting the files in order until at least 100,000 PSMs have been accepted.
The final, balanced benchmark dataset is approximately $\frac{1}{4}$ the size of the main benchmark, consisting of 779,796 PSMs from 180,250 distinct peptides.

@@

The dataset contains files resulting from various stages of the benchmark:
\begin{itemize}
\item Spectrum files in MGF format, produced by msconvert.
\item Reference proteome files in FASTA format, downloaded from Uniprot.
\item Search results files for both targets and decoys, in tab-delimited format, produced by Tide.
\item PSM-level Percolator results files for targets, in tab-delimited format.
\item Annotated MGF files for both versions of the benchmark (main and balanced).
\end{itemize}
Also included are log files for the steps of the analysis pipeline carried out using Crux \cite{park2008rapid} (Tide indexing, Tide search, and Percolator).
